Generalized spaces for constructive algebra

### 2020-04-08 paper-herrsching2019/main.tex
Proper introductions

@@

\section{Locales}

% XXX \cite{belanger-marquis:pointless} and Johnstone, vickers:locales-toposes,
% picado-pultr:frames-and-locales

@@

@@

\subsection{Geometric theories}

@@

\item \emph{The geometric theory of surjections~$f : \NN \twoheadrightarrow
\RR$.} This theory has no
sorts; no function symbols; a nullary relation symbol~$\varphi_{nx}$ for each
pair~$\langle n,x \rangle \in \NN \times \RR$ (to be read as~``$f$ maps~$n$
to~$x$''); and the following axioms:
\begin{enumerate}
\item For each~$n \in \NN$, the axiom~$\top \vdash \bigvee_{x \in \RR}
\varphi_{nx}$, expressing that~$f$ is total.
\item For each~$n \in \NN$ and each~$x,y \in \RR$, the axiom~$\varphi_{nx}
\wedge \varphi_{ny} \vdash \bigvee\{ \top \,|\, x = y \}$,
expressing that~$f$ is single-valued.\footnote{The disjunction is taken over
the set~$\{ \top \,|\, x = y \}$. This is a certain subsingleton set of
formulas; it is inhabited (by the formula~$\top$) iff~$x = y$. Hence, if~$x =
y$, this axiom reads~$\varphi_{nx}
\wedge \varphi_{ny} \vdash \top$ and could also be omitted; if~$x \neq y$,
this axiom reads~$\varphi_{nx} \wedge \varphi_{ny} \vdash \bot$. If we work in
a constructive metatheory, we cannot perform this case distinction.}
\item For each~$x \in \RR$, the axiom~$\top \vdash \bigvee_{n \in \NN}
\varphi_{nx}$, expressing that~$f$ is surjective.
\end{enumerate}

@@

\begin{remark}The semantics set up in the proof of Theorem~\ref{thm:baby-barr}
coincides with the sheaf semantics of the \emph{classifying locale} of~$\TT$
reviewed in Section~\ref{sect:presenting-frames} and
Section~\ref{sect:sheaf-semantics}. In the nonpropositional case, the semantics
coincides with the slightly more general topos semantics of the
\emph{classifying topos} of~$\TT$~\cite[Section~2]{caramello:tst}.\end{remark}

@@

% XXX Introduction...

% XXX A sheaf may well be nontrivial even though it does not have any global
% sections. This is for instance the case for the subsheaf of the sheaf of continuous
% complex-valued functions on~$\CC$, given by~$U \mapsto \{ f : U \to \CC \,|\,
% \text{$f$ is continuous and~$f(z)^2 = z$ for all~$z \in U$} \}$. This sheaf
% has lots of local sections (local square root maps), but no global one.
% This phenomenon is related to the failure of (a simple version of) the
% fundamental theorem of algebra (Exercise~\ref{ex:fta}).

Let~$X$ be a locale. A \emph{formula over an open~$U \in \O(X)$} is an
infinitary first-order formula (made up using~${=}\ {\top}\ {\wedge}\ \{\bigwedge\}\ {\bot}\ {\vee}\
{\bigvee}\ {\Rightarrow}\ {\forall}\ {\exists}$) over the signature which has one sort for
each sheaf~$F$, one constant symbol of sort~$F$ for each section~$s \in F(U)$,
one function symbol~$f : F \to G$ for each morphism of sheaves, and so on. As
usual, we define~$\neg\varphi$ to be an abbreviation of~$(\varphi \Rightarrow
\bot)$.

@@

@@

\begin{definition}\label{defn:sheaf-model}
A \emph{sheaf model} of a geometric theory~$\TT$ over a locale~$X$ consists of
\begin{enumerate}
  \item a sheaf~$\brak{S}$ on~$X$ for each sort~$S$,
  \item a morphism~$\brak{f} : \brak{S_1} \times \cdots \times \brak{S_n} \to
  \brak{T}$ of sheaves for each function symbol~$f : S_1 \times \cdots \times S_n \to T$ and
  \item a subsheaf~$\brak{R} \subseteq \brak{S_1} \times \cdots \times \brak{S_n}$
  for each relation symbol~$R \hookrightarrow S_1 \times \cdots \times S_n$
\end{enumerate}
such that the axioms of~$\TT$ are validated under the sheaf semantics.
\end{definition}

@@

A geometric theory can well have no set-based models, but might still be
consistent and allow for a diverse range of sheaf models on a host of different
locales. In fact, any propositional geometric theory always has at least one,
and very peculiar, sheaf model, namely its \emph{generic model}.

@@

@@

@@

Commutative algebra abounds with techniques to reduce given situations to
simpler ones, for instance passing to a quotient or passing to a
localization. These techniques facilitate short and elegant proofs, such as the
following. (Section~\ref{sect:algebraic-preliminaries} contains some algebraic
preliminaries.)

@@

We can also cast the contruction in logical language. This change in perspective makes it evident
that abelianization is just a special case of a more general procedure.
The free abelian group over~$G$ can be obtained as the \emph{term model} of the
theory of abelian groups extended by a constant~$x_g$ for each element~$g \in
G$ and the axioms
\begin{align*}
  \top &\vdash x_{g \circ h} = x_g \circ x_h & \text{(for each~$g,h \in G$)} \\
  \top &\vdash x_g^{-1} = x_{g^{-1}} & \text{(for each~$g \in G$)} \\
  \top &\vdash x_e = e
\end{align*}
That is, $G^\ab$ contains exactly those elements and exactly those
identifications between elements such that~$G^\ab$ is an abelian group and such
that~$G$ can be interpreted in it (that is, that there is a homomorphism~$G \to
G^\ab$), and nothing more.

@@

@@

FUTURE:
* Why the name "geometric"?

>>>>

\usepackage{hyperref}

@@

\paragraph{Locales.}
% XXX \cite{belanger-marquis:pointless} and Johnstone, vickers:locales-toposes,
% picado-pultr:frames-and-locales

@@

Section~\ref{sect:locales} explores the basics of the theory of locales, with a
focus on examples and issues for constructive mathematics.

\paragraph{Geometric theories and sheaf models.} The most sensible solution to
the task
\begin{quote}
Determine the continuous complex-valued functions~$f$ such that~$f(z)^2 = z$.
\end{quote}
is not the single set
\[ \{ f : \CC \to \CC \,|\, \text{$f$ is continuous and $f(z)^2 = z$ for all~$z \in \CC$} \}. \]
Rather, since the solvability of this functional equation varies with the
domain on which~$f$ should be defined (having no global solution defined on all
of~$\CC$ but two solutions on small disks not containing the origin), we should
consider the family
\[ (\{ f : U \to \CC \,|\, \text{$f$ is continuous and $f(z)^2 = z$ for all~$z
\in \CC$} \})_{U \in \O(\CC)} \]
of sets as the solution (or perhaps, since no solution was actually given, as
an apt description) of the task.

If we are building a prime ideal~$\ppp$ step by step, adding elements to it
when the need arises, the most sensible answer to the question
\begin{quote}
Is~$x$ contained in the prime ideal~$\ppp$?
\end{quote}
is not a single truth value, but a family of truth values, parametrized by the
possible stages of construction of~$\ppp$.

Somewhat surprisingly, these two examples can be treated by a single theory,
the \emph{theory of sheaves}. Sheaves help to organize such families in an
efficient manner and reify them as single coherent entities; the first family
is (part of) a \emph{sheaf of sets} and the second family is (part of) a
\emph{sheaf of truth values}. Sheaves will be the subject of
Sections~\ref{sect:sheaves}--\ref{sect:sheaf-semantics}.

Sheaves interact particularly well with \emph{geometric theories}, which we
review in Section~\ref{sect:geometric-theories}. Firstly, sheaves allow us to
generalize our notion of a \emph{model} of a geometric theory. For instance, a
set-based model of the geometric theory of rings is just an ordinary ring; but
now we can also consider sheaves of rings as further models.

Some geometric theories only really unfold in the world of sheaves. For
instance, the geometric theory of surjections~$\NN \twoheadrightarrow \RR$
reviewed on page~\pageref{item:theory-of-surjections} does not have any
set-based models, but it is consistent and it does have nontrivial sheaf
models. For the purpose of constructive algebra, sheaf models of some geometric
theories play a vital role, as we will discuss in Section~\ref{sect:appl}.

Secondly, geometric theories allow us to efficiently construct the spaces
(locales) on which we want to study sheaves. This is because a fundamental
feature of geometric theories (in contrast to, say, arbitrary first-order
theories) is that their models naturally organize to form a space, the
space (locale) of models reviewed in Section~\ref{sect:presenting-frames}.

% A peculiar feature of geometric theories is that (Barr's theorem)
% FUTURE

\paragraph{Applications in constructive algebra.}
Commutative algebra abounds with techniques to reduce given situations to
simpler ones, for instance passing to a quotient or passing to a
localization. These techniques facilitate short and elegant proofs, such as the
following. (Section~\ref{sect:algebraic-preliminaries} contains some algebraic
preliminaries.)

\section{Locales}
\label{sect:locales}

@@

\subsection{Geometric theories}
\label{sect:geometric-theories}

@@

\item\label{item:theory-of-surjections} \emph{The geometric theory of surjections~$f : \NN \twoheadrightarrow
\RR$.} This theory has no
sorts; no function symbols; a nullary relation symbol~$\varphi_{nx}$ for each
pair~$\langle n,x \rangle \in \NN \times \RR$ (to be read as~``$f$ maps~$n$
to~$x$''); and the following axioms:
\begin{enumerate}
\item For each~$n \in \NN$, the axiom~$\top \vdash \bigvee_{x \in \RR}
\varphi_{nx}$, expressing that~$f$ is total.
\item For each~$n \in \NN$ and each~$x,y \in \RR$, the axiom~$\varphi_{nx}
\wedge \varphi_{ny} \vdash \bigvee\{ \top \,|\, x = y \}$,
expressing that~$f$ is single-valued.\footnote{The disjunction is taken over
the set~$\{ \top \,|\, x = y \}$. This is a certain subsingleton set of
formulas; it is inhabited (by the formula~$\top$) iff~$x = y$. Hence, if~$x =
y$, this axiom reads~$\varphi_{nx}
\wedge \varphi_{ny} \vdash \top$ and could also be omitted; if~$x \neq y$,
this axiom reads~$\varphi_{nx} \wedge \varphi_{ny} \vdash \bot$. If we work in
a constructive metatheory, we cannot perform this case distinction.}
\item For each~$x \in \RR$, the axiom~$\top \vdash \bigvee_{n \in \NN}
\varphi_{nx}$, expressing that~$f$ is surjective.
\end{enumerate}

@@

\begin{remark}The semantics set up in the proof of Theorem~\ref{thm:baby-barr}
coincides with the sheaf semantics of the \emph{classifying locale} of~$\TT$
reviewed in Section~\ref{sect:presenting-frames} and
Section~\ref{sect:sheaf-semantics}. In the nonpropositional case, the semantics
coincides with the slightly more general topos semantics of the
\emph{classifying topos} of~$\TT$~\cite[Section~2]{caramello:tst}. The local
operator defined in the proof cuts out a certain Boolean subtopos of the
classifying topos.
\end{remark}

@@

\paragraph{Motivating the sheaf semantics.} Let~$F$ be a sheaf on a locale~$X$.
What is a sensible reading of the statement~``$\forall s\?F\_ (\varphi(s)
\Rightarrow \psi(s))$''? A first answer is ``for any global section~$s \in
F(X)$ such that~$\varphi(s)$,~$\psi(s)$''. However, this proposal does not do
justice to the sheaf, whose interesting behavior may well only play out with
its sections on small opens. Indeed, many important sheaves have no or few
global sections. Hence a more sophisticated answer is ``for any open~$U \in
\O(X)$ and any section~$s \in F(U)$ with~$\varphi(s)$,~$\psi(s)$'',
where~``$\varphi(s)$'' and~``$\psi(s)$'' have to be interpreted in a similar
vein.

Similarly, given a section~$s \in F(U)$, what should~``$\neg\varphi(s)$'' mean?
Since a section may well validate additional properties when we zoom in to
smaller opens, the sheaf-theoretically sensible answer is ``except for the
empty open~$V = \bot$,~$s|^U_V$ does not have property~$\varphi$ for any
open~$V \preceq U$'', where again~``$\varphi(s)$'' has to be understood in a
sensible way.

As a final motivating example, consider the statement~``$\exists s\?F\_
\varphi(s)$''. This should mean that there is an open covering~$X = \bigvee_i
U_i$ such that on each open~$U_i$, there is a section~$s \in F(U_i)$ such
that~$\varphi(s)$, properly interpreted. We should not require these individual
sections to agree on the intersections~$U_i \wedge U_j$ -- by the sheaf
property, this would amount to saying that there is a global section~$s \in
F(X)$ with~$\varphi(s)$.

The \emph{sheaf semantics} formalizes the informal language agreement just
outlined. A marvelous fact with wide-ranging applications is that this
semantics is sound with respect to first-order intuitionistic logic (and even
infinitary higher-order intuitionistic logic, and even a suitable version of
extensional type theory, if the semantics is appropriately extended).

That is, if we have established a property~$\varphi$ and if there is an
intuitionistic proof that~$\varphi$ entails a further property~$\psi$, then we
may conclude that also the interpretation of~$\psi$ with the sheaf semantics
holds.

Hence the sheaf semantics allows us to reason with sheaves in much the same way
as we reason with sets, in a simple element-based language, and to import much
of the already-existing results in constructive mathematics into the world of
sheaves. For instance, any intuitionistic theorem about rings yields a
corresponding theorem about sheaves of rings.

There is only one caveat to keep in mind: Irrespective of its status in the
metatheory, the interpretation of the law of excluded middle with the sheaf
semantics is, for most locales~$X$, false (Example~\ref{ex:negneg}). People who
are familiar with other flavors of constructive mathematics should also observe
that the interpretation of any form of the axiom of choice, even countable
choice or weak countable choice, is false for most locales~$X$. (For discrete
locales, these axioms pass from the metatheory to the universe of sheaves.)

Incidentally, that the sheaf semantics is not sound with respect to classical
logic allows us to illustrate the failure of certain classical theorems in
constructive mathematics by geometric counterexamples and thereby verify their
unprovability in intuitionistic logic.

We now begin the formal development of the sheaf semantics. More background and
examples from theoretical computer science, analysis and differential geometry
are contained in Ref.~\refcite{blechschmidt:custom-tailored}, which should be
regarded as a companion paper to this contribution.

\paragraph{Definition and basic properties of the sheaf semantics.}
Let~$X$ be a locale. A \emph{formula over an open~$U \in \O(X)$} is an
infinitary first-order formula (made up using~${=}\ {\top}\ {\wedge}\ \{\bigwedge\}\ {\bot}\ {\vee}\
{\bigvee}\ {\Rightarrow}\ {\forall}\ {\exists}$) over the signature which has one sort for
each sheaf~$F$, one constant symbol of sort~$F$ for each section~$s \in F(U)$,
one function symbol~$f : F \to G$ for each morphism of sheaves, and so on. As
usual, we define~$\neg\varphi$ to be an abbreviation of~$(\varphi \Rightarrow
\bot)$.

@@

\begin{example}\label{ex:negneg}
For instance, let~$F$ be the sheaf
\[ U \longmapsto \{ f : U \to \CC \,|\, \text{$f$ is continuous and~$f(z)^2 =
z$ for all~$z \in U$} \} \]
on~$X = \CC$. The interpretation of the statement~``$\exists s\?F\_ \top$'' is
false, as on no open which contains the origin a continuous square root
function exists), but the interpretation of~``$\neg\neg(\exists s\?F\_ \top)$''
is true: Unrolling the definition, this statement expresses that
\begin{quote}
the only open~$U \preceq X$ which has the property that \\
${\qquad}$ the only open~$V \preceq U$ which has the property that \\
${\qquad\qquad}$ locally on~$V$, there exist continuous square root functions \\
${\qquad}$ is~$V = \bot$ \\
is~$U = \bot$.
\end{quote}
\end{example}

@@

% FUTURE: Add an introductory sentence.
\paragraph{Sheaf models.}
\begin{definition}\label{defn:sheaf-model}
A \emph{sheaf model} of a geometric theory~$\TT$ over a locale~$X$ consists of
\begin{enumerate}
  \item a sheaf~$\brak{S}$ on~$X$ for each sort~$S$,
  \item a morphism~$\brak{f} : \brak{S_1} \times \cdots \times \brak{S_n} \to
  \brak{T}$ of sheaves for each function symbol~$f : S_1 \times \cdots \times S_n \to T$ and
  \item a subsheaf~$\brak{R} \subseteq \brak{S_1} \times \cdots \times \brak{S_n}$
  for each relation symbol~$R \hookrightarrow S_1 \times \cdots \times S_n$
\end{enumerate}
such that the axioms of~$\TT$ are validated under the sheaf semantics.
\end{definition}

@@

For further reference we state a proposition relating truth in a sheaf model
with truth at all points.
% FUTURE: Give better motivation

\begin{proposition}Let~$\TT$ be a geometric theory. Let~$\TT_0$ be the same
theory as~$\TT$ but without any axioms. Let~$M$ be a sheaf model
of~$\TT_0$ on a locale~$X$ (hence just a \emph{sheaf structure}) and consider
the structures~$M_x$ obtained by computing the stalks at points~$x \in \Pt(X)$.
Let~$\sigma$ be a geometric sequent.
Then:
\begin{enumerate}
\item If~$M$ validates~$\sigma$, then so does~$M_x$, for any point~$x \in \Pt(X)$.
\item The converse holds if~$X$ is spatial.
\end{enumerate}
\end{proposition}

\begin{proof}Omitted.\end{proof}

In particular, the stalks of sheaf models of a geometric theory~$\TT$ are
set-based models of~$\TT$.

\paragraph{The generic model.}
A geometric theory can well have no set-based models, but might still be
consistent and allow for a diverse range of sheaf models on a host of different
locales. In fact, any propositional geometric theory always has at least one,
and very peculiar, sheaf model, namely its \emph{generic model}.

@@

For a geometric sequent~$\sigma$ over the signature of a geometric
theory~$\TT$, the following statements are equivalent:
\begin{enumerate}
\item The generic model of~$\TT$ validates~$\sigma$.
\item All sheaf models of~$\TT$ on arbitrary locales validate~$\sigma$.
\item The theory~$\TT$ proves~$\sigma$:
\end{enumerate}
Hence we may understand the phrase~``Let~$M$ be a model of~$\TT$.'' to
implicitly refer to the generic model. The first-order properties of the
generic model need not be shared by all sheaf models and need not be provable
by~$\TT$.
% FUTURE: Explain more

@@

\begin{exercise}{The least number principle}%
\begin{alphlist}[(e)]
\item A subset~$M \subseteq X$ of a set~$X$ is \emph{detachable} iff for any
element~$x \in X$, either~$x \in M$ or~$x \not\in M$. Show that any inhabited
detachable subset~$M \subseteq \NN$ contains a minimal element.
\item Show that for any inhabited set~$M$ of natural numbers, it is \notnot the
case that there is a number~$n \in M$ such that~$n \leq m$ for all~$m \in M$.
\item Show that for every finitely generated vector space~$V$ over a field in the
sense that nonunits are zero, it is \notnot the case that~$V$ has a basis.
\item Verify that the assumption that any inhabited set of natural numbers
contains a minimal element implies the law of excluded middle.\smallskip

{\scriptsize\emph{Hint.} Consider a set of the form~$\{ n \in \NN \,|\, n = 1
\vee \varphi \}$.\par}
\item Give an explicit example of a subsheaf~$F$ of the constant
sheaf~$\underline{\NN}$ on a topological space~$X$ such that
\[ \Sh(X) \models \exists n\?F\_ \top
  \quad\text{but}\quad
  \Sh(X) \not\models \exists n\?F\_ \forall m\?F\_ n \leq m. \]
\end{alphlist}
\end{exercise}

% FUTURE: Failure of IVT

@@

We recall from the introduction one way how locales and sheaves yield new
reduction techniques for constructive algebra:
\begin{enumerate}
\item Instead of replacing a given ring with another (for instance a ring~$A$ with
one of its stalks~$A_\ppp$ at a minimal prime ideal), we replace a given ring
with a \emph{sheaf of rings}.
\item We then maintain the convenient illusion that we are working with a plain old ring
instead of a sheaf by employing the sheaf semantics.
\end{enumerate}

@@

We can also cast the contruction in logical language. This change in perspective makes it evident
that abelianization is just a special case of a more general procedure.
The free abelian group over~$G$ can be obtained as the \emph{term model} of the
theory of abelian groups extended by a constant~$x_g$ for each element~$g \in
G$ and the axioms
\begin{align*}
  \top &\vdash x_{g \circ h} = x_g \circ x_h & \text{(for each~$g,h \in G$)} \\
  \top &\vdash x_{g^{-1}} = x_g^{-1} & \text{(for each~$g \in G$)} \\
  \top &\vdash x_e = e
\end{align*}
That is, $G^\ab$ contains exactly those elements and exactly those
identifications between elements such that~$G^\ab$ is an abelian group and such
that~$G$ can be interpreted in it (that is, that there is a homomorphism~$G \to
G^\ab$), and nothing more.

@@

\begin{remark}Given a ring~$A$, there is the geometric theory of \emph{local
localizations} of~$A$. This is the theory of local rings extended by constants~$e_a$
for each element~$a \in A$ and the following axioms:
\begin{align*}
  \top &\vdash e_0 = 0 \\
  \top &\vdash e_1 = 1 \\
  \top &\vdash e_{a + b} = e_a + e_b & \text{(for each~$a,b \in A$)} \\
  \top &\vdash e_{ab} = e_a e_b & \text{(for each~$a,b \in A$)} \\
  \top &\vdash e_{-a} = -e_a & \text{(for each~$a \in A$)} \\
  \top &\vdash_{x:R} \bigvee_{a \in A} \bigvee_{s \in A} \bigl((\exists z\?R\_
  e_s z = 1) \wedge e_s x = e_a\bigr) \\
  e_a = 0 &\vdash \bigvee_{\substack{s \in A\\sa = 0}} (\exists z\?R\_ e_s z = 1) & \text{(for each~$a \in A$)}
\end{align*}
By the first five listed entries, a model~$R$ of this theory comes equipped
with a ring homomorphism~$f : A \to R$. The remaining two entries ensure that~$R$ is
a localization of~$A$, more precisely that it is the localization of~$A$
at~$f^{-1}R^\times$.

Since this theory is not propositional, Section~\ref{sect:presenting-frames}
does not apply. There is, however, still a classifying topos of this theory, and it turns
out that it coincides with the topos of sheaves over~$\Spec(A)$ -- mainly
because the theory of local localizations of~$A$ is equivalent to the theory of
prime filters of~$A$. Under this identification, the generic model of the local
localization is precisely the sheaf~$A^\sim$, and it validates first-order
sequents not expected from a local localization; as repeatedly stressed
(statement~(\ref{item:field})), it is a field, even though the theory of local
localizations does not prove the field condition.
\end{remark}
% FUTURE: Cite Nullstellensatz paper

@@

FUTURE: Why the name "geometric"?
FUTURE: Forcing
